---
title: "Introduction to R"
author: "Andi"
Last updated: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Assign days as names of poker_vector
names(poker_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign days as names of roulette_vector
names(roulette_vector)<- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

poker_vector
```


```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Select poker results for Monday, Tuesday and Wednesday
poker_start <- poker_vector[c("Monday", "Tuesday", "Wednesday")]
  
# Calculate the average of the elements in poker_start
mean(poker_start)
```

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0
  
# Print out selection_vector
selection_vector
```

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0

# Select from poker_vector these days
poker_winning_days <- poker_vector[selection_vector]
```

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on roulette?
selection_vector <- roulette_vector > 0

# Select from roulette_vector these days
roulette_winning_days <- roulette_vector[selection_vector]
```

```{r}
# Construct a matrix with 3 rows that contain the numbers 1 up to 9, filled row-wise
matrix(1:9, byrow = TRUE, nrow = 3)

# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Construct matrix
star_wars_matrix <- matrix(c(new_hope, empire_strikes, return_jedi), nrow = 3, byrow = TRUE)

# Vectors region and titles, used for naming
region <- c("US", "non-US")
titles <- c("A New Hope", "The Empire Strikes Back", "Return of the Jedi")

# Name the columns with region
colnames(star_wars_matrix)<-region

# Name the rows with titles
rownames(star_wars_matrix)<-titles

# Print out star_wars_matrix
star_wars_matrix
```

## Calculating the worldwide box office

```{r}
# Construct star_wars_matrix
box_office <- c(460.998, 314.4, 290.475, 247.900, 309.306, 165.8)
region <- c("US", "non-US")
titles <- c("A New Hope", 
                 "The Empire Strikes Back", 
                 "Return of the Jedi")
               
star_wars_matrix <- matrix(box_office, 
                      nrow = 3, byrow = TRUE,
                      dimnames = list(titles, region))

# Calculate worldwide box office figures
worldwide_vector <- rowSums(star_wars_matrix)
```

## Adding a column for the Worldwide box office

```{r}
# Construct star_wars_matrix
box_office <- c(460.998, 314.4, 290.475, 247.900, 309.306, 165.8)
region <- c("US", "non-US")
titles <- c("A New Hope", 
            "The Empire Strikes Back", 
            "Return of the Jedi")
               
star_wars_matrix <- matrix(box_office, 
                      nrow = 3, byrow = TRUE,
                      dimnames = list(titles, region))

# The worldwide box office figures
worldwide_vector <- rowSums(star_wars_matrix)

# Bind the new variable worldwide_vector as a column to star_wars_matrix
all_wars_matrix <- cbind(star_wars_matrix, worldwide_vector)
```

## Adding a row

```{r}
# Box office Star Wars sequels (in millions!)
The_Phantom_Menace <- c(474.5,  552.5)
Attack_of_the_Clones <- c(310.7,  338.7)
Revenge_of_the_Sith <- c(380.3,  468.5)

# Construct matrix
star_wars_matrix2 <- matrix(c(The_Phantom_Menace, Attack_of_the_Clones, Revenge_of_the_Sith), nrow = 3, byrow = TRUE)

# Vectors region and titles, used for naming
region <- c("US", "non-US")
titles <- c("The Phantom Menace", "Attack of the Clones", "Revenge of the Sith")

# Name the columns with region
colnames(star_wars_matrix2)<-region

# Name the rows with titles
rownames(star_wars_matrix2)<-titles

# Print out star_wars_matrix
star_wars_matrix2

# Combine both Star Wars trilogies in one matrix
all_wars_matrix <- rbind(star_wars_matrix, star_wars_matrix2)
```

## The total box office revenue for the entire saga

```{r}
# all_wars_matrix is available in your workspace
all_wars_matrix

# Total revenue for US and non-US
total_revenue_vector <- colSums(all_wars_matrix)
  
# Print out total_revenue_vector
total_revenue_vector
```

## Selection of matrix elements

```{r}
# all_wars_matrix is available in your workspace
all_wars_matrix

# Select the non-US revenue for all movies
non_us_all <- all_wars_matrix[,2]
  
# Average non-US revenue
mean(non_us_all)
  
# Select the non-US revenue for first two movies
non_us_some <- all_wars_matrix[c(1,2),2]
  
# Average non-US revenue for first two movies
mean(non_us_some)
```

## nominal categorical variable and ordinal categorical variable

```{r}
# Animals
animals_vector <- c("Elephant", "Giraffe", "Donkey", "Horse")
factor_animals_vector <- factor(animals_vector)
factor_animals_vector

# Temperature
temperature_vector <- c("High", "Low", "High","Low", "Medium")
factor_temperature_vector <- factor(temperature_vector, order = TRUE, levels = c("Low", "Medium", "High"))
factor_temperature_vector
```

## Factor levels

```{r}
# Code to build factor_survey_vector
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
factor_survey_vector

# Specify the levels of factor_survey_vector
levels(factor_survey_vector) <-c("Female", "Male")

factor_survey_vector
```


## Summarizing a factor

```{r}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")
factor_survey_vector

# Generate summary for survey_vector
summary(survey_vector)

# Generate summary for factor_survey_vector
summary(factor_survey_vector)
```

## Battle of the sexes

```{r}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")

# Male
male <- factor_survey_vector[1]

# Female
female <- factor_survey_vector[2]

# Battle of the sexes: Male 'larger' than female?
male > female
```

## Ordered factors

```{r}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")

# Convert speed_vector to ordered factor vector
factor_speed_vector <-factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Print factor_speed_vector
factor_speed_vector
summary(factor_speed_vector)
```

## Comparing ordered factors

```{r}
# Create factor_speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Factor value for second data analyst
da2 <-factor_speed_vector[2]

# Factor value for fifth data analyst
da5 <-factor_speed_vector[5]

# Is data analyst 2 faster than data analyst 5?
da2 > da5
```

## Data Frames

## Have a look at the structure

```{r}
# Investigate the structure of mtcars
str(mtcars)
```

## Creating a data frame

```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", 
          "Mars", "Jupiter", "Saturn", 
          "Uranus", "Neptune")
type <- c("Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", 
          "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 
              11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 
              0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# Create a data frame from the vectors
planets_df <-data.frame(name, type, diameter, rotation, rings)

# Check the structure of planets_df
str(planets_df)
```

## Selection of data frame elements

```{r}
# Print out diameter of Mercury (row 1, column 3)
planets_df[1, 3]

# Print out data for Mars (entire fourth row)
planets_df[4, ]

# Select first 5 values of diameter column
planets_df[1:5, "diameter"]
```

## Only planets with rings

```{r}
# Select the rings variable from planets_df
rings_vector <- planets_df$rings
  
# Print out rings_vector
rings_vector

# Select all columns for planets with rings
planets_df[rings_vector, ]
```

## Only planets with rings but shorter

Syntax: `subset(my_df, subset = some_condition)`

```{r}
subset(planets_df, subset = rings)

# Select planets with diameter < 1
subset(planets_df, subset = (diameter < 1))
```

## Sorting your data frame

```{r}
# Use order() to create positions
positions <- order(planets_df$diameter)

# Use positions to sort planets_df
planets_df[positions, ]
```


