---
title: "02 Intermediate R"
author: "Andi"
Last updated: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    keep_md: yes
---

## Relational operators

```{r}
# Comparison of logicals
TRUE == FALSE

# Comparison of numerics
-6*14 != 17 - 101

# Comparison of character strings
"useR" == "user"

# Compare a logical with a numeric
TRUE == 1

# Comparison of numerics
-6 * 5 + 2 >= -10 + 1

# Comparison of character strings
"raining" <= "raining dogs"

# Comparison of logicals
TRUE > FALSE

# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Popular days
linkedin > 15

# Quiet days
linkedin <= 5

# LinkedIn more popular than Facebook
linkedin > facebook
```

## comparing matrices

```{r}
# The social data has been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)
views <- matrix(c(linkedin, facebook), nrow = 2, byrow = TRUE)

# When does views equal 13?
views == 13

# When is views less than or equal to 14?
views <= 14
```

## Logical operators

## & and |

```{r}
# The linkedin and last variable are already defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
last <- tail(linkedin, 1)

# Is last under 5 or above 10?
last < 5 | last > 10

# Is last between 15 (exclusive) and 20 (inclusive)?
15 < last & last <= 20

# The social data (linkedin, facebook, views) has been created for you

# linkedin exceeds 10 but facebook below 10
linkedin > 10 & facebook < 10

# When were one or both visited at least 12 times?
linkedin >= 12 | facebook >= 12

# When is views between 11 (exclusive) and 14 (inclusive)?
views > 11 & views <= 14
```

## Blend it all together

```{r, eval=FALSE}
# li_df is pre-loaded in your workspace

# Select the second column, named day2, from li_df: second
second <- li_df[, "day2"]

# Build a logical vector, TRUE if value in second is extreme: extremes
extremes <- second > 25 | second <5

# Count the number of TRUEs in extremes
sum(extremes)
```

## Conditonal statements

## if statement
```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Examine the if statement for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
}

# Write the if statement for num_views
if (num_views > 15){print("You are popular!")}
```

## add an else

```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Control structure for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else {
  print("Unknown medium")
}



# Control structure for num_views
if (num_views > 15) {
  print("You're popular!")
} else {
  print("Try to be more visible!")
}
```

## further customize: else if

```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Control structure for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else if (medium == "Facebook") {
  # Add code to print correct string when condition is TRUE
  print("Showing Facebook information")
} else {
  print("Unknown medium")
}

# Control structure for num_views
if (num_views > 15) {
  print("You're popular!")
} else if (num_views <= 15 & num_views > 10) {
  # Add code to print correct string when condition is TRUE
  print("Your number of views is average")
} else {
  print("Try to be more visible!")
}
```

## Take control!

```{r}
# Variables related to your last day of recordings
li <- 15
fb <- 9

# Code the control-flow construct
if (li >= 15 & fb >= 15) {
  sms <- 2 * (li + fb)
} else if (li < 10 & fb < 10) {
  sms <- 0.5 * (li + fb)
} else {
  sms <- li + fb
}

# Print the resulting sms to the console
sms
```

## while loop

## write a while loop

```{r}
# Initialize the speed variable
speed <- 64

# Code the while loop
while (speed > 30) {
  print("Slow down!")
  speed <- speed - 7
}

# Print out the speed variable
speed
```

## Throw in more conditionals

```{r}
# Initialize the speed variable
speed <- 64

# Extend/adapt the while loop
while (speed > 30) {
  print(paste("Your speed is",speed))
  if (speed > 48) {
    print("Slow down big time!")
    speed = speed - 11
  } else {
    print("Slow down!")
    speed = speed - 6
  }
}
```

## Stop the while loop: break

```{r}
# Initialize the speed variable
speed <- 88

while (speed > 30) {
  print(paste("Your speed is", speed))
  
  # Break the while loop when speed exceeds 80
  if (speed > 80) {
  break
  }
  
  if (speed > 48) {
    print("Slow down big time!")
    speed <- speed - 11
  } else {
    print("Slow down!")
    speed <- speed - 6
  }
}
```

## Build a while loop from scratch

```{r}
# Initialize i as 1 
i <- 1

# Code the while loop
while (i <= 10) {
  print(3*i)
  if (i %% 8 == 0) {
    break
    print(3*i)
  }
  i <- i + 1
}
```

## The for loop

## Loop over a vector

```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Loop version 1
for (x in linkedin){
    print(x)
}

# Loop version 2
for (i in 1: length(linkedin)){
    print(linkedin[i])
}
```

## Loop over a list

```{r}
# The nyc list is already specified
nyc <- list(pop = 8405837, 
            boroughs = c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"), 
            capital = FALSE)

# Loop version 1
for (x in nyc){
    print(x)
}


# Loop version 2
for (i in 1: length(nyc)){
    print(nyc[[i]]) # notice the double square brackets
}
```

## Loop over a matrix

```{r}
# Define the tic-tac-toe matrix ttt 
ttt<-matrix(c("O",  NA,   "X", NA,  "O",  "O", "X",  NA,   "X" ), 
            byrow = T, nrow = 3);ttt

# define the double for loop
for (i in 1: nrow(ttt)) {
  for (j in 1: ncol(ttt)) {
    print(paste("On row", i, "and column", j, "the board contains", ttt[i,j]))
  }
}
```

## Mix it up with control flow

```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Code the for loop with conditionals
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")    
  } else {
    print("Be more visible!")
  }
  print(li)
}
```

## Next, you break it

```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Adapt/extend the for loop
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  
  # Add if statement with break
  if (li > 16){
    print("This is ridiculous, I'm outta here!")
    break
  }
  
  
  # Add if statement with next
  if (li < 5){
    print("This is too embarrassing!")
    next
  }
  

  print(li)
}

# notice that "break" and "next" have to be inside the expression of the for loop
```

## Build a for loop from scratch

This exercise will not introduce any new concepts on for loops.

We already went ahead and defined a variable rquote. This variable has been split up into a vector that contains separate letters and has been stored in a vector chars with the strsplit() function.

Can you write code that counts the number of r's that come before the first u in rquote?

```{r}
# Pre-defined variables
rquote <- "r's internals are irrefutably intriguing"
chars <- strsplit(rquote, split = "")[[1]]

# Initialize rcount
rcount <- 0

# Finish the for loop
for (char in chars) {
    if (char == "r"){
        rcount = rcount + 1
    }else if (char == "u"){
        break
    }  
  
}

# Print out rcount
print(rcount)
```

## Introduction to Functions

## Function documentation
```{r}
# Consult the documentation on the mean() function
# ?mean

# Inspect the arguments of the mean() function
args(mean)
```


## Use a function
Check the documentation on the `mean()` function again:

`?mean`

The Usage section of the documentation includes two versions of the `mean()` function. The first usage,

`mean(x, ...)`

is the most general usage of the mean function. The `'Default S3 method',` however, is:

`mean(x, trim = 0, na.rm = FALSE, ...)`

The `...` is called the ellipsis. It is a way for R to pass arguments along without the function having to name them explicitly. The ellipsis will be treated in more detail in future courses.

For the remainder of this exercise, just work with the second usage of the mean function. Notice that both trim and na.rm have default values. This makes them optional arguments.
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Calculate the mean of the sum
avg_sum <- mean(linkedin + facebook)

# Calculate the trimmed mean of the sum
avg_sum_trimmed <- mean(linkedin + facebook, trim = 0.2)

# Inspect both new variables
avg_sum
avg_sum_trimmed
```

## Nested functions

```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, NA, 17, 14)
facebook <- c(17, NA, 5, 16, 8, 13, 14)

# Calculate the mean absolute deviation
mean(abs(linkedin - facebook), na.rm = T)
```

## Writing functions

Using `return()` inside an `if` statement is a bit like using `break` inside an `if` statement. 

There are situations in which your function does not require an input. Let's say you want to write a function that gives us the random outcome of throwing a fair die:

```{r}
set.seed(528)

throw_die <- function() {
  number <- sample(1:6, size = 1)
  number
}

throw_die()
```

code a function that doesn't take any arguments

```{r}
# Define the function hello()
hello <- function(){
    print("Hi there!")
    return(TRUE)
}

# Call the function hello()
hello()
```

## A function with default arguments

```{r}
# Finish the pow_two() function
pow_two <- function(x, print_info = TRUE) {
  y <- x ^ 2
  if (print_info == TRUE) {
    print(paste(x, "to the power two equals", y))
    }
  return(y)
}

pow_two(3)

pow_two(3, print_info = FALSE)
```

## Function scoping

Whatever variable defined inside the function is not available to call from outside the function.

## A more advanced function

```{r}
# Back to the linkedin and facebook vectors 

# Define the interpret function
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Call the interpret function twice
linkedin <- linkedin[complete.cases(linkedin)]
facebook <- facebook[complete.cases(facebook)]
interpret(linkedin[1])
interpret(facebook[2])

```

## Still more advanced

```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# The interpret() can be used inside interpret_all()
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Define the interpret_all() function
# views: vector with data to interpret
# return_sum: return total number of views on popular days?
interpret_all <- function(views, return_sum = TRUE) {
  count <- 0

  for (v in views) {
    count = count + interpret(v)
  }

  if (return_sum == TRUE) {
    return(count)
  } else {
    return(NULL)
  }
}

# Call the interpret_all() function on both linkedin and facebook
interpret_all(linkedin)
interpret_all(facebook)
```

## R packages

```{r}
# look at the currently attached packages 
search()

# Load the ggplot2 package
library(ggplot2)

# Retry the qplot() function
qplot(mtcars$wt, mtcars$hp)

# Check out the currently attached packages again
search()
```

## The apply family

## lapply()

## Use lapply with a built-in R function

```{r}
# The vector pioneers has already been created for you
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)
```

## Use lapply with your own function

```{r}
# Code from previous exercise:
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Write function select_first()
select_first <- function(x) {
  x[1]
}

# Apply select_first() over split_low: names
names <- lapply(split_low, select_first)

# Write function select_second()
select_second <- function(x) {
  x[2]
}

# Apply select_second() over split_low: years
years <- lapply(split_low, select_second)
```

## lapply and anonymous functions

```{r}
# split_low has been created for you
split_low

# Transform: use anonymous function inside lapply
names <- lapply(split_low, function(x) {x[1]})

# Transform: use anonymous function inside lapply
years <- lapply(split_low, function(x) {x[2]})
```

## Use lapply with additional arguments

```{r}
# Definition of split_low
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Generic select function
select_el <- function(x, index) {
  x[index]
}

# Use lapply() twice on split_low: names and years
names <- lapply(split_low, select_el, index = 1)
years <- lapply(split_low, select_el, index = 2)
```

